smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-08-02 manuscript/manuscript.qmd
adding section on missing data patterns

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostics to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.

@@

The smdi R package was written in R language (version 4.2.1)

## Theoretical Background and Main Package Functions

For the implementation of these routine missing data diagnostics, we categorized the main functions of the `smdi` R package into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**). In this section, we will explain the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results. Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates (one-row-per-patient format). Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions consider automatically any variable in the dataframe that exhibits at least one missing value.

### Group 1 Diagnostics

@@

### Group 2 Diagnostics

@@

### Group 3 Diagnostics

@@

`r table_cap(name = "tbl2", "Patient characteristics of the exemplary cohort used to showcase the smdi package.")`

```{r}
#| echo: false
#| message: false

labels <- readxl::read_excel(
  path = here("tables", "tables_manual.xlsx"),
  sheet = "covariates"
  )

labels_list <- as.list(labels$`Variable Label`)
names(labels_list) <- labels$`Variable Name`

# set compact theme 
theme_gtsummary_compact(font_size = 11)

tbl2 <- smdi_data %>% 
  select(-c(eventtime, status)) %>% 
  mutate(exposure = ifelse(exposure == 0, "Drug A", "Drug B")) %>% 
  mutate(
    across(
      c(
        female_cat, 
        smoking_cat, 
        physical_cat, 
        alk_cat, 
        histology_cat, 
        copd_cat
        ), 
      ~ifelse(.x == "0", 0, 1)
      )
    ) %>% 
  tbl_summary(
    by = "exposure", 
    label = labels_list, 
    missing = "ifany"
    ) %>% 
  add_overall() %>% 
  add_p() %>% 
  as_flex_table() %>% 
  font(fontname = "Calibri") %>% 
  align(align = "left")

tbl2
```

`r table_cap(name = "tbl3", "Overview of covariates of de novo simulated cohort used to showcase the smdi package.")`

read_excel(
  path = here('tables', 'tables_manual.xlsx'),
  sheet = "covariates"
  ) %>% 
  select(-`Variable Name`) %>% 
  gt() %>% 
  tab_style(
    cell_text(weight = "bold"),
    locations = cells_column_labels(columns = everything())
    )
```

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "workflow.jpg"))
```

```{r}
#| label: fig-univariate
#| fig-cap: "Complete case analysis."
#| echo: false
#| eval: false
#| out-width: 110%

library(survival)
library(broom)
suppressPackageStartupMessages(library(mice))
library(survival)

# specifiy form
cox_lhs <- "survival::Surv(eventtime, status)"
cox_rhs <- paste(colnames(smdi_data %>% select(-c(eventtime, status))), collapse = " + ")
cox_form <- as.formula(paste(cox_lhs, "~", cox_rhs))

# fit univariate
cox_univariate <- survival::coxph(survival::Surv(eventtime, status) ~ exposure, 
                                  data = smdi_data) %>% 
  {. ->> cox_n} %>% 
  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
  filter(term == "exposure") %>% 
  mutate(
    method = "Univariate",
    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
    events = formatC(cox_n$nevent, big.mark = ",", format = "fg"),
    ) %>% 
  select(method, n, events, estimate, conf.low, conf.high)

# fit cox for complete case analysis
cox_cca <- survival::coxph(cox_form, data = smdi_data) %>% 
  {. ->> cox_n} %>% 
  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
  filter(term == "exposure") %>% 
  mutate(
    method = "Complete case analysis",
    n = formatC(cox_n$n, big.mark = ",", format = "fg"),
    events = formatC(cox_n$nevent, big.mark = ",", format = "fg")
    ) %>% 
  select(method, n, events, estimate, conf.low, conf.high)

# fit multiple imputation model and cox model
cox_mi <- mice::futuremice(
  data = smdi_data,
  method = "rf",
  rfPackage = "randomForest",
  m = 50,
  n.core = parallel::detectCores(),
  parallelseed = 42,
  print = FALSE
  ) %>% 
  with(expr = survival::coxph(formula(paste(format(cox_form), collapse = "")))) %>%
  mice::pool() %>%
  {. ->> cox_n} %>% 
  broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>% 
  filter(term == "exposure") %>% 
  mutate(
    method = "Multiple imputation",
    n = formatC(mean(cox_n$glanced$n), big.mark = ",", format = "fg"),
    events = formatC(mean(cox_n$glanced$nevent), big.mark = ",", format = "fg")
    ) %>% 
  select(method, n, events, estimate, conf.low, conf.high)

bind_rows(cox_univariate, cox_cca, cox_mi) %>% 
  mutate(across(where(is.numeric), ~formatC(.x, digits = 2, format = "f"))) %>% 
  gt()
```

{{< pagebreak >}}
:::

>>>>

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in studies [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostic analyses to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.

@@

The `smdi` R package was written in R language (version 4.2.1) and using the devtools [@devtools] and roxygen [@roxygen2] packages. The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command. To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.

## Main Package Functions

@fig-workflow illustrates the recommended workflow and functions that can be used to systematically approach diagnostics on partially observed confounders.

In general, the workflow comprises two broader diagnostic categories: descriptives and pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. This section covers the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results. Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions consider automatically any variable in the dataframe that exhibits at least one missing value.

### Descriptives and Pattern Diagnostics

As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates. The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations for partially observed covariates which can also be stratified by one or more grouping variables (e.g., by an exposure or outcome variable). The `smdi_vis()` function returns a corresponding bar chart plot.

[@ruddle2022]

### Inferential Three Group Diagnostics

The main functions used to investigate the potential underlying missing data mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).

#### Group 1 Diagnostics

@@

#### Group 2 Diagnostics

@@

#### Group 3 Diagnostics

@@

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "workflow.jpg"))
```
:::
